---
title: "Experiment 2: Preferred Task Delay"
author: "Przemysław Marcowski, PhD"
---

```{r setup, include=FALSE}
#| code-fold: true

library(tidyverse)
library(brms)
library(easystats)
library(patchwork)
library(see)

# Select confidence interval
ci_level <- 0.99

set.seed(42) # for reproducibility
```

## Overview

Experiment 2 tests how task difficulty and attractiveness jointly shape preferred delay. Participants reported when they would prefer to complete tasks in four conditions crossing easy versus hard with boring versus interesting.

The analysis models log-transformed delay with a Gaussian mixed model. Participant-specific intercepts account for repeated observations, and 99% intervals carry uncertainty through the model summaries, marginal means, and contrasts.

## Design and sample

The prepared data contain one row per task-delay response. The notebook reports only aggregate sample information; participant-level records are not displayed.

```{r exp2-import}
# Read prepared data from Experiment 2
exp2_delay_data <- readRDS("../data/prepared/exp2.Rds")
```

```{r exp2-data-summary}
#| echo: false

tibble::tibble(
  `Data feature` = c("Responses", "Participants", "Variables", "Experimental conditions", "Missing cells"),
  Value = c(
    format(nrow(exp2_delay_data), big.mark = ","),
    format(dplyr::n_distinct(exp2_delay_data$id), big.mark = ","),
    ncol(exp2_delay_data),
    dplyr::n_distinct(interaction(exp2_delay_data$difficulty, exp2_delay_data$attractiveness)),
    format(sum(is.na(exp2_delay_data)), big.mark = ",")
  )
) |>
  gt::gt()
```

### Sample characteristics

```{r exp2-demographics}
# Inspect sample characteristics
exp2_sample <- exp2_delay_data %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  summarize(
    n = n_distinct(id),
    mean_age = mean(age),
    sd_age = sd(age),
    n_male = n_distinct(id[gender == "Male"]),
    n_female = n_distinct(id[gender == "Female"]),
    n_other = n_distinct(id[gender == "Other"])
  )

exp2_sample %>%
  transmute(
    Participants = n,
    `Mean age` = round(mean_age, 1),
    `Age SD` = round(sd_age, 1),
    Male = n_male,
    Female = n_female,
    Other = n_other
  ) %>%
  gt::gt()
```

### Follow-up task preferences

Counts of preferred task types for immediate and delayed execution.

```{r exp2-followup}
# Count follow-up task preferences
exp2_delay_data %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, which_now, which_later) %>%
  gather(task, pref, which_now, which_later) %>%
  group_by(task, pref) %>%
  count() %>%
  ungroup() %>%
  spread(pref, n)
```

## Analysis

The Gaussian mixed model predicts log-delay from difficulty, attractiveness, and their interaction. Random intercepts represent stable between-participant differences in preferred delay.

### Model specification

```{r exp2-model-spec}
# Model formula and family
exp2_model_formula <- bf(paste0("log_t ~ difficulty * attractiveness + (1|id)"))
exp2_model_family <- brmsfamily("gaussian")
exp2_model_priors <- get_prior(exp2_model_formula, data = exp2_delay_data, family = exp2_model_family)

exp2_model_priors
```

### Model fitting

```{r exp2-model-fit}
# Fit Gaussian model
exp2_model_fit <-
  brm(
    exp2_model_formula,
    data = exp2_delay_data, family = exp2_model_family, prior = exp2_model_priors,
    iter = 4000, warmup = 2000, chains = 4, cores = 4,
    file = paste0("../output/exp2_gaussian_model_fit")
  )
```

## Results

### Model summary

```{r exp2-model-summary}
# Inspect model
summary(exp2_model_fit, prob = ci_level)
```

### Posterior description

```{r exp2-posterior}
# Describe posterior
exp2_posterior_desc <- describe_posterior(exp2_model_fit, ci = ci_level, rope_ci = ci_level)

exp2_posterior_desc
```

### Marginal means

Predicted log-delay by difficulty and attractiveness condition.

```{r exp2-means}
# Calculate posterior means
exp2_marginal_means <- estimate_means(exp2_model_fit, at = c("difficulty", "attractiveness"), ci = ci_level, rope_ci = ci_level)

exp2_marginal_means
```

### Contrasts

Pairwise differences between factor levels: attractiveness contrasts within each difficulty level, and difficulty contrasts within each attractiveness level.

#### Attractiveness contrasts

```{r exp2-contrasts-attr}
# Calculate attractiveness contrasts by difficulty level
exp2_contrasts_attr <- estimate_contrasts(exp2_model_fit, contrast = "attractiveness", by = "difficulty", ci = ci_level, rope_ci = ci_level, p_adjust = "mvt")

exp2_contrasts_attr
```

#### Difficulty contrasts

```{r exp2-contrasts-diff}
# Calculate difficulty contrasts by attractiveness level
exp2_contrasts_diff <- estimate_contrasts(exp2_model_fit, contrast = "difficulty", by = "attractiveness", ci = ci_level, rope_ci = ci_level, p_adjust = "mvt")

exp2_contrasts_diff
```

### Figure

Marginal delay preferences and pairwise contrast estimates across conditions.

#### Preferred task delay

```{r exp2-means-plot}
#| fig-show: hide
#| fig-cap: "Model-estimated log-delay by task difficulty and attractiveness. Error bars show 99% intervals."
#| fig-alt: "Line plot of estimated log-delay for easy and hard tasks, with separate series for boring and interesting tasks and 99% interval bars."

# Visualize means across difficulty and attractiveness
exp2_means_plot <-
  exp2_marginal_means %>%
  mutate(
    log_t = Median
  ) %>%
  ggplot(aes(x = difficulty, y = log_t, group = attractiveness, color = attractiveness)) +
  geom_point(size = 1) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.1, linewidth = 0.5) +
  geom_line(linewidth = 0.5) +
  labs(
    title = "Preferences in Task Delay",
    x = "Task Difficulty", y = "log(t)",
    color = "Task Appeal"
  ) +
  coord_cartesian(ylim = range(exp2_delay_data$log_t)) +
  scale_color_grey() +
  scale_fill_grey() +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1.1),
    plot.margin = unit(c(0, 3, 0, 0), "lines")
  )

exp2_means_plot
```

#### Attractiveness contrasts

```{r exp2-contrasts-attr-plot}
#| fig-show: hide
#| fig-cap: "Attractiveness contrasts within each difficulty condition. Error bars show 99% intervals."
#| fig-alt: "Horizontal contrast plot comparing boring with interesting tasks separately for easy and hard tasks. A dashed line marks no difference."

# Visualize attractiveness contrasts
exp2_contrasts_attr_plot <-
  exp2_contrasts_attr %>%
  ggplot(aes(x = difficulty, y = Median, fill = difficulty)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(color = "black", width = 0.25, linewidth = 0.5) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.1, linewidth = 0.5) +
  labs(
    title = "Differences in Task Delay\n(Preference by Difficulty)",
    x = "Task Difficulty", y = "log(t) Boring - Interesting",
    fill = "Task Difficulty"
  ) +
  guides(fill = "none") +
  coord_flip(ylim = c(-2, 2)) +
  scale_fill_manual(values = c("white", "white")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1)
  )

exp2_contrasts_attr_plot
```

#### Difficulty contrasts

```{r exp2-contrasts-diff-plot}
#| fig-show: hide
#| fig-cap: "Difficulty contrasts within each attractiveness condition. Error bars show 99% intervals."
#| fig-alt: "Horizontal contrast plot comparing easy with hard tasks separately for boring and interesting tasks. A dashed line marks no difference."

# Visualize difficulty contrasts
exp2_contrasts_diff_plot <-
  exp2_contrasts_diff %>%
  ggplot(aes(x = attractiveness, y = Median, fill = attractiveness)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(color = "black", width = 0.25, linewidth = 0.5) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.1, linewidth = 0.5) +
  labs(
    title = "Differences in Task Delay\n(Preference by Attractiveness)",
    x = "Task Attractiveness", y = "log(t) Easy - Hard",
    fill = "Attractiveness"
  ) +
  guides(fill = "none") +
  coord_flip(ylim = c(-2, 2)) +
  scale_fill_manual(values = c("white", "white")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 0.8)
  )

exp2_contrasts_diff_plot
```

#### Combined results

```{r exp2-figure}
#| fig-cap: "Preferred task delay and condition contrasts in Experiment 2. Panel a shows model-estimated log-delay; panels b and c show attractiveness and difficulty contrasts. Intervals are 99%."
#| fig-alt: "Three-panel figure showing estimated task delay, boring-minus-interesting contrasts, and easy-minus-hard contrasts across the experimental conditions."
#| fig-width: 10
#| fig-height: 4

# Create and save figure 2
fig2 <-
  exp2_means_plot + exp2_contrasts_attr_plot + exp2_contrasts_diff_plot +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a") &
    theme(
      plot.title = element_text(size = 12),
      legend.position = "bottom"
      )

print(fig2)

ggsave("../output/fig2.png", fig2, width = 10, height = 4)
```

## Interpretation

The marginal means describe expected delay in each condition. The contrasts isolate the attractiveness difference at each difficulty level and the difficulty difference at each attractiveness level. For each estimate, the posterior median gives the effect direction and magnitude; the 99% interval shows the remaining uncertainty.

## Supplement and reproducibility

The model uses a fixed seed and a cached fit under `output/`. The session record identifies the software environment used to produce the displayed summaries and figure.

### Session information

```{r session}
sessionInfo()
```
