---
title: "Experiment 3: Task Choice Across Delays"
author: "Przemysław Marcowski, PhD"
---

```{r setup, include=FALSE}
#| code-fold: true

library(tidyverse)
library(brms)
library(easystats)
library(patchwork)
library(see)

# Select confidence interval
ci_level <- 0.99
emmeans::emm_options(level = ci_level)

set.seed(42) # for reproducibility
```

## Overview

Experiment 3 tests whether the delay before a task changes its likelihood of being chosen, and whether that relationship depends on task difficulty and attractiveness. Participants made binary choices across six delay levels in four task conditions.

The analysis uses a Bernoulli logit mixed model with all interactions among difficulty, attractiveness, and log-delay. Participant-specific intercepts account for repeated choices, and 99% intervals carry uncertainty through the posterior summaries and derived estimates.

## Design and sample

The prepared data contain one row per task choice. The notebook reports only aggregate sample information; participant-level records are not displayed.

```{r exp3-import}
# Read prepared data from Experiment 3
exp3_task_data <- readRDS("../data/prepared/exp3.Rds")
```

```{r exp3-data-summary}
#| echo: false

tibble::tibble(
  `Data feature` = c("Choices", "Participants", "Variables", "Condition-by-delay cells", "Missing cells"),
  Value = c(
    format(nrow(exp3_task_data), big.mark = ","),
    format(dplyr::n_distinct(exp3_task_data$id), big.mark = ","),
    ncol(exp3_task_data),
    dplyr::n_distinct(interaction(exp3_task_data$difficulty, exp3_task_data$attractiveness, exp3_task_data$log_t)),
    format(sum(is.na(exp3_task_data)), big.mark = ",")
  )
) |>
  gt::gt()
```

### Sample characteristics

```{r exp3-demographics}
# Inspect sample characteristics
exp3_sample <- exp3_task_data %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  summarize(
    n = n_distinct(id),
    mean_age = mean(age),
    sd_age = sd(age),
    n_male = n_distinct(id[gender == "Male"]),
    n_female = n_distinct(id[gender == "Female"]),
    n_other = n_distinct(id[gender == "Other"])
  )

exp3_sample %>%
  transmute(
    Participants = n,
    `Mean age` = round(mean_age, 1),
    `Age SD` = round(sd_age, 1),
    Male = n_male,
    Female = n_female,
    Other = n_other
  ) %>%
  gt::gt()
```

### Empirical choice proportions

Observed choice proportions by difficulty, attractiveness, and delay level, with 99% confidence intervals.

```{r exp3-proportions}
# Inspect empirical choice proportions
props3 <-
  exp3_task_data %>%
  group_by(difficulty, attractiveness, log_t, chosen) %>%
  count(chosen) %>%
  ungroup() %>%
  group_by(difficulty, attractiveness, log_t) %>%
  mutate(
    prop = n / sum(n),
    ci_objs = lapply(n, function(k) prop.test(k, sum(n), conf.level = ci_level)),
    CI_high = sapply(ci_objs, function(x) x$conf.int[2]),
    CI_low = sapply(ci_objs, function(x) x$conf.int[1])
  ) %>%
  filter(chosen > 0)

props3
```

## Analysis

The Bernoulli logit mixed model predicts task choice from difficulty, attractiveness, log-delay, and their interactions. Random intercepts represent stable between-participant differences in choice propensity.

### Model specification

```{r exp3-model-spec}
# Model formula and family
exp3_model_formula <- bf(paste0("chosen ~ difficulty*attractiveness*log_t + (1|id)"))
exp3_model_family <- brmsfamily("bernoulli", "logit")
exp3_model_priors <- get_prior(exp3_model_formula, data = exp3_task_data, family = exp3_model_family)

exp3_model_priors
```

### Model fitting

```{r exp3-model-fit}
# Fit Bernoulli model with logit link
exp3_model_fit <-
  brm(
    exp3_model_formula,
    data = exp3_task_data, family = exp3_model_family, prior = exp3_model_priors,
    iter = 4000, warmup = 2000, chains = 4, cores = 4,
    file = "../output/exp3_bernoulli_model_fit"
  )
```

## Results

### Model summary

```{r exp3-model-summary}
# Inspect model
summary(exp3_model_fit, prob = ci_level)
```

### Posterior description

```{r exp3-posterior}
# Describe posterior
exp3_posterior_desc <- bayestestR::describe_posterior(exp3_model_fit, ci = ci_level, rope_ci = ci_level)

exp3_posterior_desc
```

### Marginal means and contrasts

Predicted choice probabilities across the delay range and overall, with pairwise contrasts between attractiveness levels within each difficulty condition.

#### Marginal means by delay

```{r exp3-means-time}
# Calculate marginal means across time
exp3_marginal_means_time <- estimate_means(exp3_model_fit, by = c("difficulty", "attractiveness", "log_t"), length = 100, ci = ci_level, rope_ci = ci_level)

exp3_marginal_means_time
```

#### Estimated marginal means

```{r exp3-emmeans}
# Calculate marginal means and contrasts
exp3_emmeans <- emmeans::emmeans(exp3_model_fit, pairwise ~ attractiveness | difficulty, regrid = "response", infer = TRUE, adjust = "mvt")
exp3_marginal_means <- as.data.frame(summary(exp3_emmeans$emmeans, level = ci_level))

exp3_marginal_means
```

#### Pairwise contrasts

```{r exp3-contrasts}
# Pairwise contrasts
exp3_contrasts <- as.data.frame(summary(exp3_emmeans$contrasts), level = ci_level)
exp3_contrasts$sig <- with(exp3_contrasts, ifelse(lower.HPD <= 0 & upper.HPD >= 0, "", "*"))

exp3_contrasts
```

### Slope estimates

Standardized delay slopes by difficulty and attractiveness, quantifying how choice probability changes with delay in each condition.

```{r exp3-slopes}
# Calculate trends for task delay across conditions
exp3_slopes <- standardize(estimate_slopes(exp3_model_fit, trend = "log_t", by = c("difficulty", "attractiveness"), ci = ci_level, rope_ci = ci_level))

exp3_slopes

exp3_slopes_contrasts <- estimate_contrasts(exp3_model_fit, trend = "log_t", by = "attractiveness", ci = ci_level)
print(exp3_slopes_contrasts)

exp3_slopes_contrasts
```

### Figure

Predicted acceptance curves, marginal means, pairwise contrasts, and delay slope estimates.

#### Task acceptance by delay

```{r exp3-means-time-plot}
#| fig-show: hide
#| fig-cap: "Model-estimated task acceptance across log-delay, separated by attractiveness and difficulty. Bands show 99% intervals."
#| fig-alt: "Two faceted curves show task acceptance across log-delay for easy and hard tasks, with uncertainty bands and a horizontal reference at one-half."

# Visualize means across task delay
exp3_means_time_plot <-
  exp3_marginal_means_time %>%
  ggplot(aes(x = log_t, y = Median, group = difficulty, color = difficulty, fill = difficulty)) +
  geom_line(linewidth = 0.5) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high), color = "transparent", alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = max(exp3_task_data$log_t) / 2, linetype = "dashed") +
  labs(
    title = "Task Acceptance by Delay",
    x = "log(t)", y = "P(Choose)",
    color = "Task Difficulty",
    fill = "Task Difficulty"
  ) +
  coord_cartesian(
    xlim = range(exp3_task_data$log_t),
    ylim = c(0, 1)
  ) +
  scale_color_grey(start = 0.1, end = 0.4) +
  scale_fill_grey(start = 0.1, end = 0.4) +
  theme_modern() +
  facet_wrap(~attractiveness, ncol = 2, scales = "free_x") +
  theme(
    aspect.ratio = 1,
    legend.position = "bottom",
    legend.justification = c(0.5, 0.5),
    plot.title = element_text(hjust = 0.55),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

exp3_means_time_plot
```

#### Marginal task acceptance

```{r exp3-means-plot}
#| fig-show: hide
#| fig-cap: "Marginal task-acceptance probability by difficulty and attractiveness. Error bars show 99% intervals."
#| fig-alt: "Line plot of marginal task-acceptance probability for easy and hard tasks, with separate boring and interesting series and 99% interval bars."

# Visualize means across conditions
exp3_means_plot <-
  exp3_marginal_means %>%
  ggplot(aes(x = difficulty, y = response, group = attractiveness, color = attractiveness)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.05, linewidth = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  labs(
    title = "Marginal Task Acceptance",
    x = "Task Difficulty", y = "P(Choose)",
    color = "Task Appeal"
  ) +
  guides(color = "none") +
  coord_cartesian(ylim = c(0, 1)) +
  scale_color_grey() +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 0.9),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

exp3_means_plot
```

#### Differences in task acceptance

```{r exp3-contrasts-plot}
#| fig-show: hide
#| fig-cap: "Attractiveness contrasts in task acceptance within each difficulty condition. Error bars show 99% intervals."
#| fig-alt: "Horizontal contrast plot comparing boring with interesting tasks separately for easy and hard tasks. A dashed line marks no difference."

# Visualize attractiveness contrasts
exp3_contrasts_plot <-
  exp3_contrasts %>%
  ggplot(aes(x = difficulty, y = estimate, fill = difficulty)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(color = "black", width = 0.25, linewidth = 0.5) +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.1, linewidth = 0.5) +
  labs(
    title = "Differences in Task Acceptance",
    x = "Task Difficulty", y = "P(Choose) Boring - Interesting",
    fill = "Task Difficulty"
  ) +
  guides(fill = "none") +
  coord_flip(ylim = c(-1, 1)) +
  scale_fill_manual(values = c("white", "white")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1.1),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

exp3_contrasts_plot
```

#### Effects of task delay on preference

```{r exp3-slopes-plot}
#| fig-show: hide
#| fig-cap: "Standardized delay slopes by task difficulty and attractiveness. Error bars show 99% intervals."
#| fig-alt: "Horizontal point-and-interval plot of standardized delay slopes for boring and interesting tasks at each difficulty level. A dashed line marks zero."

# Visualize slope estimates
exp3_slopes_plot <-
  exp3_slopes %>%
  ggplot(aes(x = difficulty, y = Median, color = attractiveness)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(position = position_dodge(width = 0.5), size = 1) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), position = position_dodge(width = 0.5), width = 0.2, linewidth = 0.5) +
  coord_flip(ylim = c(-0.05, 0.05)) +
  labs(
    title = "Effects of Task Delay on Preference",
    x = "Task Difficulty", y = bquote(beta),
    color = "Task Appeal"
  ) +
  scale_y_continuous(labels = function(x) sprintf("%.2f", x)) +
  guides(color = "none") +
  scale_color_grey() +
  scale_fill_grey() +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1.5)
  )

exp3_slopes_plot
```

#### Combined results

```{r exp3-figure}
#| fig-cap: "Task acceptance across delay in Experiment 3. Panel a shows predicted choice curves; panels b through d show delay slopes, marginal acceptance, and attractiveness contrasts. Intervals are 99%."
#| fig-alt: "Four-panel figure showing predicted task acceptance over delay, standardized delay slopes, marginal acceptance by condition, and boring-minus-interesting contrasts."
#| fig-width: 10
#| fig-height: 8

# Create and save figure 3
fig3 <-
  ((exp3_means_time_plot) / (exp3_slopes_plot + exp3_means_plot + exp3_contrasts_plot)) +
    plot_annotation(tag_levels = "a") &
    theme(
      plot.title = element_text(size = 12),
      legend.position = "bottom"
      )

print(fig3)

ggsave("../output/fig3.png", fig3, width = 10, height = 8)
```

## Interpretation

The acceptance curves show how choice probability changes across the delay range. Marginal means average over delay, attractiveness contrasts compare boring with interesting tasks within each difficulty level, and standardized slopes summarize the delay relationship in each condition. For each estimate, the posterior interval shows the uncertainty around its direction and magnitude.

## Supplement and reproducibility

The model uses a fixed seed and a cached fit under `output/`. The session record identifies the software environment used to produce the displayed summaries and figure.

### Session information

```{r session}
sessionInfo()
```
