---
title: "Data Analysis for the Temporal Task Scheduling Project (Experiment 3)"
author: "Przemyslaw Marcowski, PhD"
date: "today"
---

```{r setup, message=FALSE, warning=FALSE}
#| code-fold: true

library(tidyverse)
library(brms)
library(easystats)
library(patchwork)
library(see)

# Select confidence interval
ci_level <- 0.99
emmeans::emm_options(level = ci_level)

set.seed(42) # for reproducibility

# Prepare data for each experiment
# source("../R/prepare_data.R") # Optional: prepares to `data/prepared/`
```

# Data Import

Prepared data from Experiment 3, where participants made binary task choices across six delay levels under conditions varying in difficulty and attractiveness.

```{r exp3-import}
# Read prepared data from Experiment 3
exp3_task_data <- readRDS("../data/prepared/exp3.Rds")

exp3_task_data
```

# Sample Characteristics

```{r exp3-demographics}
# Inspect sample characteristics
exp3_task_data %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  summarize(
    n = n_distinct(id),
    mean_age = mean(age),
    sd_age = sd(age),
    n_male = n_distinct(id[gender == "Male"]),
    n_female = n_distinct(id[gender == "Female"])
  )
```

# Empirical Choice Proportions

Observed choice proportions by difficulty, attractiveness, and delay level, with 99% confidence intervals.

```{r exp3-proportions}
# Inspect empirical choice proportions
props3 <-
  exp3_task_data %>%
  group_by(difficulty, attractiveness, log_t, chosen) %>%
  count(chosen) %>%
  ungroup() %>%
  group_by(difficulty, attractiveness, log_t) %>%
  mutate(
    prop = n / sum(n),
    ci_objs = lapply(n, function(k) prop.test(k, sum(n), conf.level = ci_level)),
    CI_high = sapply(ci_objs, function(x) x$conf.int[2]),
    CI_low = sapply(ci_objs, function(x) x$conf.int[1])
  ) %>%
  filter(chosen > 0)

props3
```

# Bernoulli Model

Bernoulli logit mixed model predicting task choice from difficulty, attractiveness, log-delay, and all interactions, with random intercepts by participant.

## Model Specification

```{r exp3-model-spec}
# Model formula and family
exp3_model_formula <- bf(paste0("chosen ~ difficulty*attractiveness*log_t + (1|id)"))
exp3_model_family <- brmsfamily("bernoulli", "logit")
exp3_model_priors <- get_prior(exp3_model_formula, data = exp3_task_data, family = exp3_model_family)

exp3_model_priors
```

## Model Fitting

```{r exp3-model-fit}
# Fit Bernoulli model with logit link
exp3_model_fit <-
  brm(
    exp3_model_formula,
    data = exp3_task_data, family = exp3_model_family, prior = exp3_model_priors,
    iter = 4000, warmup = 2000, chains = 4, cores = 4,
    file = "../output/exp3_bernoulli_model_fit"
  )
```

## Model Summary

```{r exp3-model-summary}
# Inspect model
summary(exp3_model_fit, prob = ci_level)
```

## Posterior Description

```{r exp3-posterior}
# Describe posterior
exp3_posterior_desc <- bayestestR::describe_posterior(exp3_model_fit, ci = ci_level, rope_ci = ci_level)

exp3_posterior_desc
```

## Save Coefficients

```{r exp3-save-coefficients}
# Save model coefficients (Optional: saves to `output/`)
# write_csv(exp3_posterior_desc, "../output/exp3_model_coefficients.csv")
```

# Marginal Means and Contrasts

Predicted choice probabilities across the delay range and overall, with pairwise contrasts between attractiveness levels within each difficulty condition.

## Marginal Means by Time

```{r exp3-means-time}
# Calculate marginal means across time
exp3_marginal_means_time <- estimate_means(exp3_model_fit, by = c("difficulty", "attractiveness", "log_t"), length = 100, ci = ci_level, rope_ci = ci_level)

exp3_marginal_means_time
```

## Estimated Marginal Means

```{r exp3-emmeans}
# Calculate marginal means and contrasts
exp3_emmeans <- emmeans::emmeans(exp3_model_fit, pairwise ~ attractiveness | difficulty, regrid = "response", infer = TRUE, adjust = "mvt")
exp3_marginal_means <- as.data.frame(summary(exp3_emmeans$emmeans, level = ci_level))

exp3_marginal_means
```

## Save Marginal Means

```{r exp3-save-means}
# Save marginal means (Optional: saves to `output/`)
# write_csv(exp3_marginal_means, "../output/exp3_marginal_means.csv")
```

## Pairwise Contrasts

```{r exp3-contrasts}
# Pairwise contrasts
exp3_contrasts <- as.data.frame(summary(exp3_emmeans$contrasts), level = ci_level)
exp3_contrasts$sig <- with(exp3_contrasts, ifelse(lower.HPD <= 0 & upper.HPD >= 0, "", "*"))

exp3_contrasts
```

## Save Contrasts

```{r exp3-save-contrasts}
# Save contrasts (Optional: saves to `output/`)
# write_csv(exp3_contrasts, "../output/exp3_pairwise_contrasts.csv")
```

# Slope Estimates

Standardized delay slopes by difficulty and attractiveness, quantifying how choice probability changes with delay in each condition.

```{r exp3-slopes}
# Calculate trends for task delay across conditions
exp3_slopes <- standardize(estimate_slopes(exp3_model_fit, trend = "log_t", by = c("difficulty", "attractiveness"), ci = ci_level, rope_ci = ci_level))

exp3_slopes

exp3_slopes_contrasts <- estimate_contrasts(exp3_model_fit, trend = "log_t", by = "attractiveness", ci = ci_level)
print(exp3_slopes_contrasts)

exp3_slopes_contrasts
```

## Save Slopes

```{r exp3-save-slopes}
# Save slope estimates (Optional: saves to `output/`)
# write_csv(exp3_slopes, "../output/exp3_slope_estimates.csv")
```

# Visualization

Predicted acceptance curves, marginal means, pairwise contrasts, and delay slope estimates.

## Task Acceptance by Delay

```{r exp3-means-time-plot}
# Visualize means across task delay
exp3_means_time_plot <-
  exp3_marginal_means_time %>%
  ggplot(aes(x = log_t, y = Median, group = difficulty, color = difficulty, fill = difficulty)) +
  geom_line(linewidth = 0.5) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high), color = "transparent", alpha = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = max(exp3_task_data$log_t) / 2, linetype = "dashed") +
  labs(
    title = "Task Acceptance by Delay",
    x = "log(t)", y = "P(Choose)",
    color = "Task Difficulty",
    fill = "Task Difficulty"
  ) +
  coord_cartesian(
    xlim = range(exp3_task_data$log_t),
    ylim = c(0, 1)
  ) +
  scale_color_grey(start = 0.1, end = 0.4) +
  scale_fill_grey(start = 0.1, end = 0.4) +
  theme_modern() +
  facet_wrap(~attractiveness, ncol = 2, scales = "free_x") +
  theme(
    aspect.ratio = 1,
    legend.position = "bottom",
    legend.justification = c(0.5, 0.5),
    plot.title = element_text(hjust = 0.55),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

exp3_means_time_plot
```

## Marginal Task Acceptance

```{r exp3-means-plot}
# Visualize means across conditions
exp3_means_plot <-
  exp3_marginal_means %>%
  ggplot(aes(x = difficulty, y = response, group = attractiveness, color = attractiveness)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1) +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.05, linewidth = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  labs(
    title = "Marginal Task Acceptance",
    x = "Task Difficulty", y = "P(Choose)",
    color = "Task Appeal"
  ) +
  guides(color = "none") +
  coord_cartesian(ylim = c(0, 1)) +
  scale_color_grey() +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 0.9),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

exp3_means_plot
```

## Differences in Task Acceptance

```{r exp3-contrasts-plot}
# Visualize attractiveness contrasts
exp3_contrasts_plot <-
  exp3_contrasts %>%
  ggplot(aes(x = difficulty, y = estimate, fill = difficulty)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(color = "black", width = 0.25, linewidth = 0.5) +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.1, linewidth = 0.5) +
  labs(
    title = "Differences in Task Acceptance",
    x = "Task Difficulty", y = "P(Choose) Boring - Interesting",
    fill = "Task Difficulty"
  ) +
  guides(fill = "none") +
  coord_flip(ylim = c(-1, 1)) +
  scale_fill_manual(values = c("white", "white")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1.1),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

exp3_contrasts_plot
```

## Effects of Task Delay on Preference

```{r exp3-slopes-plot}
# Visualize slope estimates
exp3_slopes_plot <-
  exp3_slopes %>%
  ggplot(aes(x = difficulty, y = Median, color = attractiveness)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(position = position_dodge(width = 0.5), size = 1) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), position = position_dodge(width = 0.5), width = 0.2, linewidth = 0.5) +
  coord_flip(ylim = c(-0.05, 0.05)) +
  labs(
    title = "Effects of Task Delay on Preference",
    x = "Task Difficulty", y = bquote(beta),
    color = "Task Appeal"
  ) +
  scale_y_continuous(labels = function(x) sprintf("%.2f", x)) +
  guides(color = "none") +
  scale_color_grey() +
  scale_fill_grey() +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1.5)
  )

exp3_slopes_plot
```

## Combined Figure

```{r exp3-figure}
# Create and save figure 3
fig3 <-
  ((exp3_means_time_plot) / (exp3_slopes_plot + exp3_means_plot + exp3_contrasts_plot)) +
    # plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a") &
    theme(
      plot.title = element_text(size = 12),
      legend.position = "bottom"
      )

print(fig3)

ggsave("../output/fig3.png", fig3, width = 10, height = 8)
```

# Session

```{r session}
sessionInfo()
```
