---
title: "Data Analysis for the Temporal Task Scheduling Project (Experiment 2)"
author: "Przemyslaw Marcowski, PhD"
date: "today"
---

```{r setup, message=FALSE, warning=FALSE}
#| code-fold: true

library(tidyverse)
library(brms)
library(easystats)
library(patchwork)
library(see)

# Select confidence interval
ci_level <- 0.99

set.seed(42) # for reproducibility

# Prepare data for each experiment
# source("../R/prepare_data.R") # Optional: prepares to `data/prepared/`
```

# Data Import

Prepared data from Experiment 2, where participants indicated preferred task delay across four conditions defined by crossing difficulty (easy, hard) and attractiveness (boring, interesting).

```{r exp2-import}
# Read prepared data from Experiment 2
exp2_delay_data <- readRDS("../data/prepared/exp2.Rds")

exp2_delay_data
```

# Sample Characteristics

```{r exp2-demographics}
# Inspect sample characteristics
exp2_delay_data %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  summarize(
    n = n_distinct(id),
    mean_age = mean(age),
    sd_age = sd(age),
    n_male = n_distinct(id[gender == "Male"]),
    n_female = n_distinct(id[gender == "Female"])
  )
```

# Follow-up Task Preferences

Counts of preferred task types for immediate and delayed execution.

```{r exp2-followup}
# Count follow-up task preferences
exp2_delay_data %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup() %>%
  select(id, which_now, which_later) %>%
  gather(task, pref, which_now, which_later) %>%
  group_by(task, pref) %>%
  count() %>%
  ungroup() %>%
  spread(pref, n)
```

# Gaussian Model

Gaussian mixed model predicting log-transformed delay from difficulty, attractiveness, and their interaction, with random intercepts by participant.

## Model Specification

```{r exp2-model-spec}
# Model formula and family
exp2_model_formula <- bf(paste0("log_t ~ difficulty * attractiveness + (1|id)"))
exp2_model_family <- brmsfamily("gaussian")
exp2_model_priors <- get_prior(exp2_model_formula, data = exp2_delay_data, family = exp2_model_family)

exp2_model_priors
```

## Model Fitting

```{r exp2-model-fit}
# Fit Gaussian model
exp2_model_fit <-
  brm(
    exp2_model_formula,
    data = exp2_delay_data, family = exp2_model_family, prior = exp2_model_priors,
    iter = 4000, warmup = 2000, chains = 4, cores = 4,
    file = paste0("../output/exp2_gaussian_model_fit")
  )
```

## Model Summary

```{r exp2-model-summary}
# Inspect model
summary(exp2_model_fit, prob = ci_level)
```

## Posterior Description

```{r exp2-posterior}
# Describe posterior
exp2_posterior_desc <- describe_posterior(exp2_model_fit, ci = ci_level, rope_ci = ci_level)

exp2_posterior_desc
```

## Save Coefficients

```{r exp2-save-coefficients}
# Save model coefficients (Optional: saves to `output/`)
# write_csv(exp2_posterior_desc, "../output/exp2_model_coefficients.csv")
```

# Marginal Means

Predicted log-delay by difficulty and attractiveness condition.

```{r exp2-means}
# Calculate posterior means
exp2_marginal_means <- estimate_means(exp2_model_fit, at = c("difficulty", "attractiveness"), ci = ci_level, rope_ci = ci_level)

exp2_marginal_means
```

## Save Marginal Means

```{r exp2-save-means}
# Save marginal means (Optional: saves to `output/`)
# write_csv(exp2_marginal_means, "../output/exp2_marginal_means.csv")
```

# Contrasts

Pairwise differences between factor levels: attractiveness contrasts within each difficulty level, and difficulty contrasts within each attractiveness level.

## Attractiveness Contrasts

```{r exp2-contrasts-attr}
# Calculate attractiveness contrasts by difficulty level
exp2_contrasts_attr <- estimate_contrasts(exp2_model_fit, contrast = "attractiveness", by = "difficulty", ci = ci_level, rope_ci = ci_level, p_adjust = "mvt")

exp2_contrasts_attr
```

### Save Attractiveness Contrasts

```{r exp2-save-contrasts-attr}
# Save attractiveness contrasts (Optional: saves to `output/`)
# write_csv(exp2_contrasts_attr, "../output/exp2_attractiveness_contrasts.csv")
```

## Difficulty Contrasts

```{r exp2-contrasts-diff}
# Calculate difficulty contrasts by attractiveness level
exp2_contrasts_diff <- estimate_contrasts(exp2_model_fit, contrast = "difficulty", by = "attractiveness", ci = ci_level, rope_ci = ci_level, p_adjust = "mvt")

exp2_contrasts_diff
```

### Save Difficulty Contrasts

```{r exp2-save-contrasts-diff}
# Save difficulty contrasts (Optional: saves to `output/`)
# write_csv(exp2_contrasts_diff, "../output/exp2_difficulty_contrasts.csv")
```

# Visualization

Marginal delay preferences and pairwise contrast estimates across conditions.

## Preferences in Task Delay

```{r exp2-means-plot}
# Visualize means across difficulty and attractiveness
exp2_means_plot <-
  exp2_marginal_means %>%
  mutate(
    log_t = Median
  ) %>%
  ggplot(aes(x = difficulty, y = log_t, group = attractiveness, color = attractiveness)) +
  geom_point(size = 1) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.1, linewidth = 0.5) +
  geom_line(linewidth = 0.5) +
  labs(
    title = "Preferences in Task Delay",
    x = "Task Difficulty", y = "log(t)",
    color = "Task Appeal"
  ) +
  coord_cartesian(ylim = range(exp2_delay_data$log_t)) +
  scale_color_grey() +
  scale_fill_grey() +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1.1),
    plot.margin = unit(c(0, 3, 0, 0), "lines")
  )

exp2_means_plot
```

## Attractiveness Contrasts Plot

```{r exp2-contrasts-attr-plot}
# Visualize attractiveness contrasts
exp2_contrasts_attr_plot <-
  exp2_contrasts_attr %>%
  ggplot(aes(x = difficulty, y = Median, fill = difficulty)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(color = "black", width = 0.25, linewidth = 0.5) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.1, linewidth = 0.5) +
  labs(
    title = "Differences in Task Delay\n(Preference by Difficulty)",
    x = "Task Difficulty", y = "log(t) Boring - Interesting",
    fill = "Task Difficulty"
  ) +
  guides(fill = "none") +
  coord_flip(ylim = c(-2, 2)) +
  scale_fill_manual(values = c("white", "white")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1)
  )

exp2_contrasts_attr_plot
```

## Difficulty Contrasts Plot

```{r exp2-contrasts-diff-plot}
# Visualize difficulty contrasts
exp2_contrasts_diff_plot <-
  exp2_contrasts_diff %>%
  ggplot(aes(x = attractiveness, y = Median, fill = attractiveness)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(color = "black", width = 0.25, linewidth = 0.5) +
  geom_errorbar(aes(ymin = CI_low, ymax = CI_high), width = 0.1, linewidth = 0.5) +
  labs(
    title = "Differences in Task Delay\n(Preference by Attractiveness)",
    x = "Task Attractiveness", y = "log(t) Easy - Hard",
    fill = "Attractiveness"
  ) +
  guides(fill = "none") +
  coord_flip(ylim = c(-2, 2)) +
  scale_fill_manual(values = c("white", "white")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 0.8)
  )

exp2_contrasts_diff_plot
```

## Combined Figure

```{r exp2-figure}
# Create and save figure 2
fig2 <-
  exp2_means_plot + exp2_contrasts_attr_plot + exp2_contrasts_diff_plot +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a") &
    theme(
      plot.title = element_text(size = 12),
      legend.position = "bottom"
      )

print(fig2)

ggsave("../output/fig2.png", fig2, width = 10, height = 4)
```

# Session

```{r session}
sessionInfo()
```
