---
title: "Data Analysis for the Temporal Task Scheduling Project (Experiment 1)"
author: "Przemyslaw Marcowski, PhD"
date: "today"
---

```{r setup, message=FALSE, warning=FALSE}
#| code-fold: true

library(tidyverse)
library(brms)
library(easystats)
library(patchwork)
library(see)

# Select confidence interval
ci_level <- 0.99
emmeans::emm_options(level = ci_level)

set.seed(42) # for reproducibility

# Prepare data for each experiment
# source("../R/prepare_data.R") # Optional: prepares to `data/prepared/`
```

# Data Import

Prepared data from Experiment 1, where participants chose between easy/boring and hard/interesting tasks under immediate and future timing conditions.

```{r exp1-import}
# Read prepared data from Experiment 1
exp1_freq_data <- readRDS("../data/prepared/exp1.Rds")

exp1_freq_data
```

# Demographics

## Sample Size

```{r exp1-sample-size}
# Number of participants
n_distinct(exp1_freq_data$id)
```

## Age Statistics

```{r exp1-age}
# Age statistics
mean(exp1_freq_data$age)
sd(exp1_freq_data$age)
```

## Demographics by Sex

```{r exp1-demographics}
# Demographics by sex
exp1_freq_data %>%
  group_by(sex) %>%
  summarize(
    n = n_distinct(id),
    mean_age = mean(age),
    sd_age = sd(age)
  )
```

# Preference Proportions

Task choice counts and proportions by timing condition.

```{r exp1-proportions}
# Calculate preference proportions
exp1_prop_table <-
  exp1_freq_data %>%
  count(time, choice) %>%
  group_by(time) %>%
  mutate(
    total = sum(n),
    prop = n / total
  ) %>%
  arrange(time)

exp1_prop_table
```

# Binomial Model

Binomial logit model predicting choice counts from task type, timing condition, and their interaction.

## Model Specification

```{r exp1-model-spec}
# Model formula and family
exp1_model_formula <- bf(paste0("n | trials(total) ~ choice * time"))
exp1_model_family <- brmsfamily("binomial", "logit")
exp1_model_priors <- get_prior(exp1_model_formula, data = exp1_prop_table, family = exp1_model_family)

exp1_model_priors
```

## Model Fitting

```{r exp1-model-fit}
# Fit a binomial model with identity link
exp1_model_fit <-
  brm(
    exp1_model_formula,
    data = exp1_prop_table, family = exp1_model_family, prior = exp1_model_priors,
    iter = 4000, warmup = 2000, chains = 4, cores = 4,
    file = "../output/exp1_binomial_model_fit"
  )
```

## Model Summary

```{r exp1-model-summary}
# Inspect model
summary(exp1_model_fit, prob = ci_level)
```

## Posterior Description

```{r exp1-posterior}
# Describe posterior
exp1_posterior_desc <- bayestestR::describe_posterior(exp1_model_fit, ci = ci_level, rope_ci = ci_level)

exp1_posterior_desc
```

## Save Coefficients

```{r exp1-save-coefficients}
# Save model coefficients (Optional: saves to `output/`)
# write_csv(exp1_posterior_desc, "../output/exp1_model_coefficients.csv")
```

# Marginal Means and Contrasts

Estimated marginal means and pairwise contrasts on the response scale, comparing choice probabilities between task types within each timing condition.

## Estimated Marginal Means

```{r exp1-emmeans}
# Calculate marginal means and contrasts
exp1_emmeans <- emmeans::emmeans(exp1_model_fit, pairwise ~ choice | time, regrid = "response", infer = TRUE, adjust = "mvt")
exp1_marginal_means <- as.data.frame(summary(exp1_emmeans$emmeans, level = ci_level))

exp1_marginal_means
```

## Save Marginal Means

```{r exp1-save-means}
# Save marginal means (Optional: saves to `output/`)
# write_csv(exp1_marginal_means, "../output/exp1_marginal_means.csv")
```

## Pairwise Contrasts

```{r exp1-contrasts}
# Pairwise contrasts
exp1_contrasts <- as.data.frame(summary(exp1_emmeans$contrasts, level = ci_level))
exp1_contrasts$sig <- with(exp1_contrasts, ifelse(lower.HPD <= 0 & upper.HPD >= 0, "", "*"))

exp1_contrasts
```

## Save Contrasts

```{r exp1-save-contrasts}
# Save contrasts (Optional: saves to `output/`)
# write_csv(exp1_contrasts, "../output/exp1_pairwise_contrasts.csv")
```

# Visualization

Marginal choice probabilities and pairwise differences across timing conditions.

## Marginal Task Preference

```{r exp1-means-plot}
# Visualize means across conditions
exp1_means_plot <-
  exp1_marginal_means %>%
  ggplot(aes(x = time, y = prob, group = choice, color = choice)) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.5) +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.1, linewidth = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  labs(
    title = "Marginal Task Preference",
    x = "Task Delay", y = "P(Choose)",
    color = "Task Type"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_color_grey(labels = c("Easy" = "Easy and Boring", "Hard" = "Hard and Interesting")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 0.9),
    plot.margin = unit(c(0, 3, 0, 0), "lines"),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

exp1_means_plot
```

## Differences in Task Preference

```{r exp1-contrasts-plot}
# Visualize contrasts
exp1_contrasts_plot <-
  exp1_contrasts %>%
  ggplot(aes(x = time, y = estimate, fill = time)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(color = "black", width = 0.25, linewidth = 0.5) +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.1, linewidth = 0.5) +
  labs(
    title = "Differences in Task Preference",
    x = "Task Delay", y = "P(Choose) Easy and Boring - Hard and Interesting",
    fill = "Task Delay"
  ) +
  guides(fill = "none") +
  coord_flip(ylim = c(-1, 1)) +
  scale_fill_manual(values = c("white", "white")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1.1),
    plot.margin = unit(c(0, 2, 0, 0), "lines"),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

exp1_contrasts_plot
```

## Combined Figure

```{r exp1-figure}
# Create and save figure 1
fig1 <-
  exp1_means_plot + exp1_contrasts_plot +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a") &
    theme(
      plot.title = element_text(size = 12),
      legend.position = "bottom"
      )

print(fig1)

ggsave("../output/fig1.png", fig1, width = 10, height = 4)
```

# Session

```{r session}
sessionInfo()
```
