---
title: "Pilot Study: Difficulty and Interest Ratings"
author: "Przemysław Marcowski, PhD"
---

```{r setup, include=FALSE}
#| code-fold: true

library(tidyverse)
library(BayesFactor)
library(bayestestR)

# Select confidence interval
ci_level <- 0.99
ci_pct <- sprintf("%d%%", round(ci_level * 100))

set.seed(42) # for reproducibility
```

## Overview

The pilot study compared how the same tasks were rated in experimental and survey contexts. Bayesian paired-sample tests quantify context differences in perceived difficulty and interest.

## Design and sample

Each participant rated task difficulty and interest in both contexts. The prepared dataset contains one row per participant; no participant-level records are displayed.

```{r pilot-import}
# Read pilot data
pilot_data <- readRDS("../data/prepared/pilot.Rds")
```

```{r pilot-sample}
#| tbl-cap: "Pilot sample"

tibble(
  Metric = c("Participants", "Mean age", "Age SD"),
  Value = c(
    n_distinct(pilot_data$id),
    mean(pilot_data$age, na.rm = TRUE),
    sd(pilot_data$age, na.rm = TRUE)
  )
) %>%
  gt::gt() %>%
  gt::fmt_number(columns = Value, decimals = 1)
```

### Descriptive ratings

Mean, standard deviation, and sample size for each rating measure across both contexts.

```{r pilot-descriptives}
# Calculate descriptive statistics
descriptive_stats <- pilot_data %>%
  summarize(
    experiment_hard_mean = mean(experiment_hard, na.rm = TRUE),
    experiment_hard_sd = sd(experiment_hard, na.rm = TRUE),
    experiment_hard_n = sum(!is.na(experiment_hard)),
    survey_hard_mean = mean(survey_hard, na.rm = TRUE),
    survey_hard_sd = sd(survey_hard, na.rm = TRUE),
    survey_hard_n = sum(!is.na(survey_hard)),
    experiment_interesting_mean = mean(experiment_interesting, na.rm = TRUE),
    experiment_interesting_sd = sd(experiment_interesting, na.rm = TRUE),
    experiment_interesting_n = sum(!is.na(experiment_interesting)),
    survey_interesting_mean = mean(survey_interesting, na.rm = TRUE),
    survey_interesting_sd = sd(survey_interesting, na.rm = TRUE),
    survey_interesting_n = sum(!is.na(survey_interesting))
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("task", "measure", "stat"),
    names_pattern = "(.*)_(.*)_(.*)",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>%
  mutate(
    task = str_to_title(task),
    measure = case_when(
      measure == "hard" ~ "Difficulty",
      measure == "interesting" ~ "Interest",
      TRUE ~ measure
    )
  )

descriptive_stats %>%
  select(Context = task, Measure = measure, Mean = mean, SD = sd, N = n) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(Mean, SD), decimals = 2)
```

## Analysis

Bayesian paired t-tests comparing experimental vs. survey ratings for difficulty and interest, using a medium Cauchy prior (r = 0.707) on the effect size.

### Task difficulty

```{r pilot-difficulty-test}
# Bayesian paired t-test for task difficulty ratings
difficulty_bf <- ttestBF(
  x = pilot_data$experiment_hard,
  y = pilot_data$survey_hard,
  paired = TRUE,
  rscale = 0.707
)

difficulty_bf
```

#### Posterior samples

```{r pilot-difficulty-posterior}
# Extract posterior samples
difficulty_posterior <- posterior(difficulty_bf, iterations = 10000)
difficulty_effect <- difficulty_posterior[, "delta"]
```

#### Posterior summary

```{r pilot-difficulty-summary}
# Calculate summary statistics for task difficulty ratings
difficulty_hdi <- hdi(difficulty_effect, ci = ci_level)
difficulty_mean <- mean(difficulty_effect)
difficulty_bf_value <- extractBF(difficulty_bf)$bf

cat("Mean effect (delta):", difficulty_mean, "\n")
cat(ci_pct, "HDI:", difficulty_hdi$CI_low, "-", difficulty_hdi$CI_high, "\n")
cat("Bayes Factor:", difficulty_bf_value, "\n")
```

### Task interest

```{r pilot-interest-test}
# Bayesian paired t-test for appeal ratings
interest_bf <- ttestBF(
  x = pilot_data$experiment_interesting,
  y = pilot_data$survey_interesting,
  paired = TRUE,
  rscale = 0.707
)

interest_bf
```

#### Posterior samples

```{r pilot-interest-posterior}
# Extract posterior samples
interest_posterior <- posterior(interest_bf, iterations = 10000)
interest_effect <- interest_posterior[, "delta"]
```

#### Posterior summary

```{r pilot-interest-summary}
# Calculate summary statistics for appeal ratings
interest_hdi <- hdi(interest_effect, ci = ci_level)
interest_mean <- mean(interest_effect)
interest_bf_value <- extractBF(interest_bf)$bf

cat("Mean effect (delta):", interest_mean, "\n")
cat(ci_pct, "HDI:", interest_hdi$CI_low, "-", interest_hdi$CI_high, "\n")
cat("Bayes Factor:", interest_bf_value, "\n")
```

## Results

Posterior mean effect sizes (Cohen's *d*), 99% HDIs, and Bayes Factors for both comparisons.

```{r pilot-summary}
# Set up data for plotting
effect_data <- data.frame(
  Measure = rep(c("Difficulty", "Interest"), each = length(difficulty_effect)),
  Effect = c(difficulty_effect, interest_effect)
)

# Create summary table for the Bayesian analysis
pilot_summary <- tibble(
  Measure = c("Difficulty", "Interest"),
  Comparison = c("Experiment vs. Survey", "Experiment vs. Survey"),
  Post_Mean_d = c(difficulty_mean, interest_mean),
  HDI_low = c(difficulty_hdi$CI_low, interest_hdi$CI_low),
  HDI_high = c(difficulty_hdi$CI_high, interest_hdi$CI_high),
  Bayes_Factor = c(difficulty_bf_value, interest_bf_value),
  Log_BF = c(log(difficulty_bf_value), log(interest_bf_value))
)

pilot_summary %>%
  select(
    summary_measure = Measure,
    summary_comparison = Comparison,
    `Posterior mean (d)` = Post_Mean_d,
    `99% HDI, lower` = HDI_low,
    `99% HDI, upper` = HDI_high,
    `Bayes factor` = Bayes_Factor,
    `Log Bayes factor` = Log_BF
  ) %>%
  gt::gt() %>%
  gt::cols_label(
    summary_measure = "Measure",
    summary_comparison = "Comparison"
  ) %>%
  gt::fmt_number(
    columns = c(
      `Posterior mean (d)`,
      `99% HDI, lower`,
      `99% HDI, upper`,
      `Bayes factor`,
      `Log Bayes factor`
    ),
    decimals = 2
  )
```

## Interpretation

Positive effects indicate higher ratings in the experimental context; negative effects indicate higher ratings in the survey context. The posterior means and 99% HDIs describe the magnitude and uncertainty of each difference, while the Bayes factors compare the paired-difference model with the null.

## Supplement and reproducibility

The tests use a Cauchy prior with scale 0.707, 10,000 posterior draws per comparison, a 99% interval, and a fixed random seed.

```{r session}
sessionInfo()
```
