---
title: "Experiment 1: Task Choice by Timing"
author: "Przemysław Marcowski, PhD"
---

```{r setup, include=FALSE}
#| code-fold: true

library(tidyverse)
library(brms)
library(easystats)
library(patchwork)
library(see)

# Select confidence interval
ci_level <- 0.99
emmeans::emm_options(level = ci_level)

set.seed(42) # for reproducibility
```

## Overview

Experiment 1 tested whether task timing changes preferences between an easy, boring task and a hard, interesting task. An aggregated binomial-logit model estimates choice probabilities by task type and timing condition.

## Design and sample

Participants chose between the two task options under tomorrow and future timing conditions. Participant records are aggregated to task-by-timing counts for modeling; no participant-level records are displayed.

```{r exp1-import}
# Read prepared data from Experiment 1
exp1_freq_data <- readRDS("../data/prepared/exp1.Rds")
```

### Sample size

```{r exp1-sample-size}
#| tbl-cap: "Experiment 1 sample size"

tibble(
  Metric = "Participants",
  Value = n_distinct(exp1_freq_data$id)
) %>%
  gt::gt() %>%
  gt::fmt_number(columns = Value, decimals = 0)
```

### Age

```{r exp1-age}
#| tbl-cap: "Participant age"

tibble(
  Statistic = c("Mean", "SD"),
  Age = c(mean(exp1_freq_data$age), sd(exp1_freq_data$age))
) %>%
  gt::gt() %>%
  gt::fmt_number(columns = Age, decimals = 1)
```

### Demographics by sex

```{r exp1-demographics}
#| tbl-cap: "Sample composition by reported sex"

# Demographics by sex
exp1_freq_data %>%
  group_by(sex) %>%
  summarize(
    n = n_distinct(id),
    mean_age = mean(age),
    sd_age = sd(age)
  ) %>%
  rename(
    demographic_sex = sex,
    demographic_n = n,
    demographic_mean_age = mean_age,
    demographic_sd_age = sd_age
  ) %>%
  gt::gt() %>%
  gt::cols_label(
    demographic_sex = "Reported sex",
    demographic_n = "Participants",
    demographic_mean_age = "Mean age",
    demographic_sd_age = "Age SD"
  ) %>%
  gt::fmt_number(
    columns = c(demographic_mean_age, demographic_sd_age),
    decimals = 1
  )
```

## Analysis

### Observed choice proportions

Task choice counts and proportions by timing condition.

```{r exp1-proportions}
# Calculate preference proportions
exp1_prop_table <-
  exp1_freq_data %>%
  count(time, choice) %>%
  group_by(time) %>%
  mutate(
    total = sum(n),
    prop = n / total
  ) %>%
  arrange(time)

exp1_prop_table %>%
  rename(
    timing_condition = time,
    task_choice = choice,
    choice_count = n,
    condition_total = total,
    choice_proportion = prop
  ) %>%
  gt::gt() %>%
  gt::cols_label(
    timing_condition = "Timing condition",
    task_choice = "Task choice",
    choice_count = "Choices",
    condition_total = "Condition total",
    choice_proportion = "Proportion"
  ) %>%
  gt::fmt_percent(columns = choice_proportion, decimals = 1)
```

### Aggregated binomial-logit model

Binomial logit model predicting choice counts from task type, timing condition, and their interaction.

#### Model specification

```{r exp1-model-spec}
# Model formula and family
exp1_model_formula <- bf(paste0("n | trials(total) ~ choice * time"))
exp1_model_family <- brmsfamily("binomial", "logit")
exp1_model_priors <- get_prior(exp1_model_formula, data = exp1_prop_table, family = exp1_model_family)

exp1_model_priors
```

#### Model fitting

```{r exp1-model-fit}
# Fit a binomial model with a logit link
exp1_model_fit <-
  brm(
    exp1_model_formula,
    data = exp1_prop_table, family = exp1_model_family, prior = exp1_model_priors,
    iter = 4000, warmup = 2000, chains = 4, cores = 4,
    file = "../output/exp1_binomial_model_fit"
  )
```

## Results

### Model summary

```{r exp1-model-summary}
# Inspect model
summary(exp1_model_fit, prob = ci_level)
```

### Posterior description

```{r exp1-posterior}
# Describe posterior
exp1_posterior_desc <- bayestestR::describe_posterior(exp1_model_fit, ci = ci_level, rope_ci = ci_level)

exp1_posterior_desc
```

### Marginal means and contrasts

Estimated marginal means and pairwise contrasts on the response scale, comparing choice probabilities between task types within each timing condition.

#### Estimated marginal means

```{r exp1-emmeans}
# Calculate marginal means and contrasts
exp1_emmeans <- emmeans::emmeans(exp1_model_fit, pairwise ~ choice | time, regrid = "response", infer = TRUE, adjust = "mvt")
exp1_marginal_means <- as.data.frame(summary(exp1_emmeans$emmeans, level = ci_level))

exp1_marginal_means %>%
  rename(
    marginal_choice = choice,
    marginal_time = time,
    marginal_probability = prob,
    marginal_lower = lower.HPD,
    marginal_upper = upper.HPD
  ) %>%
  gt::gt() %>%
  gt::cols_label(
    marginal_choice = "Task choice",
    marginal_time = "Timing condition",
    marginal_probability = "Probability",
    marginal_lower = "99% HPD, lower",
    marginal_upper = "99% HPD, upper"
  ) %>%
  gt::fmt_number(where(is.numeric), decimals = 3)
```

#### Pairwise contrasts

```{r exp1-contrasts}
# Pairwise contrasts
exp1_contrasts <- as.data.frame(summary(exp1_emmeans$contrasts, level = ci_level))
exp1_contrasts$sig <- with(exp1_contrasts, ifelse(lower.HPD <= 0 & upper.HPD >= 0, "", "*"))

exp1_contrasts %>%
  rename(
    contrast_label = contrast,
    contrast_time = time,
    contrast_estimate = estimate,
    contrast_lower = lower.HPD,
    contrast_upper = upper.HPD,
    contrast_significance = sig
  ) %>%
  gt::gt() %>%
  gt::cols_label(
    contrast_label = "Contrast",
    contrast_time = "Timing condition",
    contrast_estimate = "Estimate",
    contrast_lower = "99% HPD, lower",
    contrast_upper = "99% HPD, upper",
    contrast_significance = "Interval excludes zero"
  ) %>%
  gt::fmt_number(where(is.numeric), decimals = 3)
```

### Choice probabilities

Marginal choice probabilities and pairwise differences across timing conditions.

#### Marginal task preference

```{r exp1-means-plot}
#| fig-show: hide

# Visualize means across conditions
exp1_means_plot <-
  exp1_marginal_means %>%
  ggplot(aes(x = time, y = prob, group = choice, color = choice)) +
  geom_point(size = 1) +
  geom_line(linewidth = 0.5) +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.1, linewidth = 0.5) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  labs(
    title = "Marginal Task Preference",
    x = "Task Delay", y = "P(Choose)",
    color = "Task Type"
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_color_grey(labels = c("Easy" = "Easy and Boring", "Hard" = "Hard and Interesting")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 0.9),
    plot.margin = unit(c(0, 3, 0, 0), "lines"),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

```

#### Differences in task preference

```{r exp1-contrasts-plot}
#| fig-show: hide

# Visualize contrasts
exp1_contrasts_plot <-
  exp1_contrasts %>%
  ggplot(aes(x = time, y = estimate, fill = time)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_col(color = "black", width = 0.25, linewidth = 0.5) +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.1, linewidth = 0.5) +
  labs(
    title = "Differences in Task Preference",
    x = "Task Delay", y = "P(Choose) Easy and Boring - Hard and Interesting",
    fill = "Task Delay"
  ) +
  guides(fill = "none") +
  coord_flip(ylim = c(-1, 1)) +
  scale_fill_manual(values = c("white", "white")) +
  theme_modern() +
  theme(
    aspect.ratio = 1,
    plot.title = element_text(hjust = 1.1),
    plot.margin = unit(c(0, 2, 0, 0), "lines"),
    strip.background = element_rect(color = "transparent", fill = "transparent")
  )

```

#### Combined estimates

```{r exp1-figure}
#| fig-cap: "Marginal task-choice probabilities and within-timing contrasts in Experiment 1. Intervals are 99% highest posterior density intervals."
#| fig-alt: "Two-panel plot. The left panel shows marginal choice probabilities for easy and hard tasks under tomorrow and future timing. The right panel shows the easy-minus-hard probability contrast in each timing condition."
#| fig-width: 10
#| fig-height: 4

# Create and save figure 1
fig1 <-
  exp1_means_plot + exp1_contrasts_plot +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a") &
    theme(
      plot.title = element_text(size = 12),
      legend.position = "bottom"
      )

print(fig1)

ggsave("../output/fig1.png", fig1, width = 10, height = 4)
```

## Interpretation

The marginal means express each task's estimated choice probability. Within each timing condition, the contrast is easy and boring minus hard and interesting: positive estimates favor the easy task, and negative estimates favor the hard task. The 99% intervals show the uncertainty in both quantities.

## Supplement and reproducibility

The model uses the observed condition totals, a binomial likelihood with a logit link, default priors returned by `brms`, four chains of 4,000 iterations with 2,000 warmup iterations, 99% intervals, and a fixed random seed.

```{r session}
sessionInfo()
```
